### 2024-04-01 other/datasheet/datasheet_fire_incidents.qmd
update datasheet and references

---
title: "Datasheet for OpenDataToronto's Fire Incidents"
author: 
  - Jimmy Luc
thanks: "Code and data are available at: [https://github.com/rex009x/fire-incidents-toronto](https://github.com/rex009x/fire-incidents-toronto)"
date: today
date-format: long
abstract: "Work in progress."
format: pdf
number-sections: true
---

Extract of the questions from @gebru2021datasheets.

**Motivation**

1. *For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.*
    - TBD
2. *Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?*
    - TBD
3. *Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.*
    - TBD
4. *Any other comments?*
    - TBD

**Composition**

1. *What do the instances that comprise the dataset represent (for example, documents, photos, people, countries)? Are there multiple types of instances (for example, movies, users, and ratings; people and interactions between them; nodes and edges)? Please provide a description.*
	- TBD
2. *How many instances are there in total (of each type, if appropriate)?*
	- TBD
3. *Does the dataset contain all possible instances or is it a sample (not necessarily random) of instances from a larger set? If the dataset is a sample, then what is the larger set? Is the sample representative of the larger set (for example, geographic coverage)? If so, please describe how this representativeness was validated/verified. If it is not representative of the larger set, please describe why not (for example, to cover a more diverse range of instances, because instances were withheld or unavailable).*
	- TBD
4. *What data does each instance consist of? "Raw" data (for example, unprocessed text or images) or features? In either case, please provide a description.*
	- TBD
5. *Is there a label or target associated with each instance? If so, please provide a description.*
	- TBD
6. *Is any information missing from individual instances? If so, please provide a description, explaining why this information is missing (for example, because it was unavailable). This does not include intentionally removed information, but might include, for example, redacted text.*
	- TBD
7. *Are relationships between individual instances made explicit (for example, users' movie ratings, social network links)? If so, please describe how these relationships are made explicit.*
	- TBD
8. *Are there recommended data splits (for example, training, development/validation, testing)? If so, please provide a description of these splits, explaining the rationale behind them.*
	- TBD
9. *Are there any errors, sources of noise, or redundancies in the dataset? If so, please provide a description.*
	- TBD
10. *Is the dataset self-contained, or does it link to or otherwise rely on external resources (for example, websites, tweets, other datasets)? If it links to or relies on external resources, a) are there guarantees that they will exist, and remain constant, over time; b) are there official archival versions of the complete dataset (that is, including the external resources as they existed at the time the dataset was created); c) are there any restrictions (for example, licenses, fees) associated with any of the external resources that might apply to a dataset consumer? Please provide descriptions of all external resources and any restrictions associated with them, as well as links or other access points, as appropriate.*
	- TBD
11. *Does the dataset contain data that might be considered confidential (for example, data that is protected by legal privilege or by doctor-patient confidentiality, data that includes the content of individuals' non-public communications)? If so, please provide a description.*
	- TBD
12. *Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety? If so, please describe why.*
	- TBD
13. *Does the dataset identify any sub-populations (for example, by age, gender)? If so, please describe how these subpopulations are identified and provide a description of their respective distributions within the dataset.*
	- TBD
14. *Is it possible to identify individuals (that is, one or more natural persons), either directly or indirectly (that is, in combination with other data) from the dataset? If so, please describe how.*
	- TBD
15. *Does the dataset contain data that might be considered sensitive in any way (for example, data that reveals race or ethnic origins, sexual orientations, religious beliefs, political opinions or union memberships, or locations; financial or health data; biometric or genetic data; forms of government identification, such as social security numbers; criminal history)? If so, please provide a description.*
	- TBD
16. *Any other comments?*
	- TBD

**Collection process**

1. *How was the data associated with each instance acquired? Was the data directly observable (for example, raw text, movie ratings), reported by subjects (for example, survey responses), or indirectly inferred/derived from other data (for example, part-of-speech tags, model-based guesses for age or language)? If the data was reported by subjects or indirectly inferred/derived from other data, was the data validated/verified? If so, please describe how.*
	- TBD
2. *What mechanisms or procedures were used to collect the data (for example, hardware apparatuses or sensors, manual human curation, software programs, software APIs)? How were these mechanisms or procedures validated?*
	- TBD
3. *If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?*
	- TBD
4. *Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?*
	- TBD
5. *Over what timeframe was the data collected? Does this timeframe match the creation timeframe of the data associated with the instances (for example, recent crawl of old news articles)? If not, please describe the timeframe in which the data associated with the instances was created.*
	- TBD
6. *Were any ethical review processes conducted (for example, by an institutional review board)? If so, please provide a description of these review processes, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- TBD
7. *Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?*
	- TBD
8. *Were the individuals in question notified about the data collection? If so, please describe (or show with screenshots or other information) how notice was provided, and provide a link or other access point to, or otherwise reproduce, the exact language of the notification itself.*
	- TBD
9. *Did the individuals in question consent to the collection and use of their data? If so, please describe (or show with screenshots or other information) how consent was requested and provided, and provide a link or other access point to, or otherwise reproduce, the exact language to which the individuals consented.*
	- TBD
10. *If consent was obtained, were the consenting individuals provided with a mechanism to revoke their consent in the future or for certain uses? If so, please provide a description, as well as a link or other access point to the mechanism (if appropriate).*
	- TBD
11. *Has an analysis of the potential impact of the dataset and its use on data subjects (for example, a data protection impact analysis) been conducted? If so, please provide a description of this analysis, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- TBD
12. *Any other comments?*
	- TBD

**Preprocessing/cleaning/labeling**

1. *Was any preprocessing/cleaning/labeling of the data done (for example, discretization or bucketing, tokenization, part-of-speech tagging, SIFT feature extraction, removal of instances, processing of missing values)? If so, please provide a description. If not, you may skip the remaining questions in this section.*
	- TBD
2. *Was the "raw" data saved in addition to the preprocessed/cleaned/labeled data (for example, to support unanticipated future uses)? If so, please provide a link or other access point to the "raw" data.*
	- TBD
3. *Is the software that was used to preprocess/clean/label the data available? If so, please provide a link or other access point.*
	- TBD
4. *Any other comments?*
	- TBD

**Uses**

1. *Has the dataset been used for any tasks already? If so, please provide a description.*
	- TBD
2. *Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.*
	- TBD
3. *What (other) tasks could the dataset be used for?*
	- TBD
4. *Is there anything about the composition of the dataset or the way it was collected and preprocessed/cleaned/labeled that might impact future uses? For example, is there anything that a dataset consumer might need to know to avoid uses that could result in unfair treatment of individuals or groups (for example, stereotyping, quality of service issues) or other risks or harms (for example, legal risks, financial harms)? If so, please provide a description. Is there anything a dataset consumer could do to mitigate these risks or harms?*
	- TBD
5. *Are there tasks for which the dataset should not be used? If so, please provide a description.*
	- TBD
6. *Any other comments?*
	- TBD

**Distribution**

1. *Will the dataset be distributed to third parties outside of the entity (for example, company, institution, organization) on behalf of which the dataset was created? If so, please provide a description.*
	- TBD
2. *How will the dataset be distributed (for example, tarball on website, API, GitHub)? Does the dataset have a digital object identifier (DOI)?*
	- TBD
3. *When will the dataset be distributed?*
	- TBD
4. *Will the dataset be distributed under a copyright or other intellectual property (IP) license, and/or under applicable terms of use (ToU)? If so, please describe this license and/ or ToU, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms or ToU, as well as any fees associated with these restrictions.*
	- TBD
5. *Have any third parties imposed IP-based or other restrictions on the data associated with the instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.*
	- TBD
6. *Do any export controls or other regulatory restrictions apply to the dataset or to individual instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any supporting documentation.*
	- TBD
7. *Any other comments?*
	- TBD

**Maintenance**

1. *Who will be supporting/hosting/maintaining the dataset?*
	- TBD
2. *How can the owner/curator/manager of the dataset be contacted (for example, email address)?*
	- TBD
3. *Is there an erratum? If so, please provide a link or other access point.*
	- TBD
4. *Will the dataset be updated (for example, to correct labeling errors, add new instances, delete instances)? If so, please describe how often, by whom, and how updates will be communicated to dataset consumers (for example, mailing list, GitHub)?*
	- TBD
5. *If the dataset relates to people, are there applicable limits on the retention of the data associated with the instances (for example, were the individuals in question told that their data would be retained for a fixed period of time and then deleted)? If so, please describe these limits and explain how they will be enforced.*
	- TBD
6. *Will older versions of the dataset continue to be supported/hosted/maintained? If so, please describe how. If not, please describe how its obsolescence will be communicated to dataset consumers.*
	- TBD
7. *If others want to extend/augment/build on/contribute to the dataset, is there a mechanism for them to do so? If so, please provide a description. Will these contributions be validated/verified? If so, please describe how. If not, why not? Is there a process for communicating/distributing these contributions to dataset consumers? If so, please provide a description.*
	- TBD
8. *Any other comments?*
	- TBD

\newpage

# References

### 2024-04-02 scripts/00-simulate_data.R
updated scripts and added temporary data files

@@ -1,19 +1,15 @@
 #### Preamble ####
-# Purpose: Simulates... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Simulates an analysis dataset about fire incident occurrences in Toronto.
+# Author: Jimmy Luc
+# Date: 1 April 2024
+# Contact: jimmy.luc@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
 
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
 
 #### Simulate data ####
-# [...ADD CODE HERE...]
+# TODO set seed for simulation
 
-
-
+# TODO create a simulation dataset with variables of interest following the proper distribution

### 2024-04-02 scripts/01-download_data.R

@@ -1,26 +1,31 @@
 #### Preamble ####
-# Purpose: Downloads and saves the data from [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Downloads and saves the data from the opendatatoronto R package
+# Author: Jimmy Luc
+# Date: 1 April 2024
+# Contact: jimmy.luc@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
 
 
 #### Workspace setup ####
 library(opendatatoronto)
 library(tidyverse)
-# [...UPDATE THIS...]
 
 #### Download data ####
-# [...ADD CODE HERE TO DOWNLOAD...]
+# retrieve package id
+dataset <-
+  search_packages("Fire Incidents")
 
+# retrieve resource id
+dataset <-
+  list_package_resources(dataset$id) |> first()
+
+# retrieve dataset
+raw_fire_incidents_data <- get_resource(dataset$id)
 
 
 #### Save data ####
-# [...UPDATE THIS...]
-# change the_raw_data to whatever name you assigned when you downloaded it.
-write_csv(the_raw_data, "inputs/data/raw_data.csv") 
-
-         
+# write raw data as csv
+write_csv(
+  raw_fire_incidents_data,
+  "data/raw_data/raw_fire_incidents.csv"
+)

### 2024-04-02 scripts/02-data_cleaning.R

@@ -1,44 +1,47 @@
 #### Preamble ####
-# Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 6 April 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Cleans the raw fire incidents data
+# Author: Jimmy Luc
+# Date: 1 April 2024
+# Contact: jimmy.luc@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: Execute the 01-download_data.R script before running this file
 
 #### Workspace setup ####
 library(tidyverse)
+library(janitor)
+library(here)
 
 #### Clean data ####
-raw_data <- read_csv("inputs/data/plane_data.csv")
+# load in raw data
+raw_fire_incidents_data <-
+  read_csv(
+    file=here("data/raw_data/raw_fire_incidents.csv")
+  )
 
-cleaned_data <-
-  raw_data |>
-  janitor::clean_names() |>
-  select(wing_width_mm, wing_length_mm, flying_time_sec_first_timer) |>
-  filter(wing_width_mm != "caw") |>
-  mutate(
-    flying_time_sec_first_timer = if_else(flying_time_sec_first_timer == "1,35",
-                                   "1.35",
-                                   flying_time_sec_first_timer)
-  ) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "490",
-                                 "49",
-                                 wing_width_mm)) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "6",
-                                 "60",
-                                 wing_width_mm)) |>
-  mutate(
-    wing_width_mm = as.numeric(wing_width_mm),
-    wing_length_mm = as.numeric(wing_length_mm),
-    flying_time_sec_first_timer = as.numeric(flying_time_sec_first_timer)
-  ) |>
-  rename(flying_time = flying_time_sec_first_timer,
-         width = wing_width_mm,
-         length = wing_length_mm
-         ) |> 
-  tidyr::drop_na()
+# clean the names and select column of interest
+cleaned_fire_incidents_data <-
+  clean_names(raw_fire_incidents_data) |>
+  select(
[…]

### 2024-04-02 scripts/03-test_data.R

@@ -1,15 +1,16 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Tests the cleaned data types and values
+# Author: Jimmy Luc
+# Date: 1 April 2024
+# Contact: jimmy.luc@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: Execute the 01-download_data.R and 02-data_cleaning.R scripts before running this file
 
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
 
 #### Test data ####
+# TODO test data types for variables of interest
+
+# TODO test data values for variables of interest

### 2024-04-02 scripts/04-model.R

@@ -1,11 +1,10 @@
 #### Preamble ####
-# Purpose: Models... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Models the Estimated Dollars Loss
+# Author: Jimmy Luc
+# Date: 1 April 2024
+# Contact: jimmy.luc@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: Execute the 01-download_data.R and 02-data_cleaning.R scripts before running this file
 
 
 #### Workspace setup ####
@@ -13,25 +12,11 @@ library(tidyverse)
 library(rstanarm)
 
 #### Read data ####
-analysis_data <- read_csv("data/analysis_data/analysis_data.csv")
+analysis_data <- read_csv("data/analysis_data/cleaned_fire_incidents.csv")
 
 ### Model data ####
-first_model <-
-  stan_glm(
-    formula = flying_time ~ length + width,
-    data = analysis_data,
-    family = gaussian(),
-    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_aux = exponential(rate = 1, autoscale = TRUE),
-    seed = 853
-  )
+# TODO complete the modeling for capture the estimated dollars loss
 
 
 #### Save model ####
-saveRDS(
-  first_model,
-  file = "models/first_model.rds"
-)
-
-
+# TODO save the model to "/models" as .rds

### 2024-04-02 paper/paper.qmd
updated introduction and data sections

---
title: "Evaluating Toronto's Fire Incidents to Extrapolate Future Occurence Data"
author: 
  - Jimmy Luc
thanks: "Code and data are available at: [https://github.com/rex009x/fire-incidents-toronto](https://github.com/rex009x/fire-incidents-toronto)."
date: today
date-format: long
abstract: "This paper uses publicly available data from [Toronto Open Data](https://open.toronto.ca/) to predict the estimated dollar loss per fire incident in the Toronto area by property use. Further analysis of Toronto Fire Services' response time, area of origin, and various fire prevention measures as considering factors in the estimation. The results reveal a positive correlation between estimated dollar loss and possible causes of the incident, namely, mechanical/electrical failures yielding the highest loss. These findings bring urgency to fire safety as they highlight the importance of proper education, maintenance, and safety inspections. This analysis encourages homeowners, businesses, and property managers to take initiative in order to ensure the well-being of others."
format:
  pdf:
    fig-pos: "H"
toc: true
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(dplyr)
library(ggplot2)
library(here)
```

# Introduction

Fires pose a significant threat to both residential and commercial properties, often resulting in substantial financial losses and, more importantly, endangering lives. Developing a better understanding of the factors that influence the amount in damages per fire incident is crucial for implementing effective fire prevention measures to minimize the impact of such incidents. In this analysis, we delve into the extensive dataset provided by Toronto Open Data [@citeTorontoFireIncidents] which captures fire incidents within the Toronto area as defined by Ontario Fire Marshal (OFM) from 2011-2022. Through this dataset, we aim to predict the estimated dollar loss per fire incident across different property uses and identify the factors contributing to these losses.

The motivation behind this study stems from the necessity to comprehend the underlying patterns and determinants of fire incident costs. By dissecting the data provided by Toronto Open Data, we seek to not only predict the estimated dollar loss but also to identify significant factors influencing these losses. Our analysis extends beyond mere prediction, delving into factors such as Toronto Fire Services' response time, area of origin, and the presence and functionality of fire alarms. These additional variables are essential in understanding the dynamics of fire incidents and formulating informed strategies for fire prevention and mitigation.

Despite the abundance of data available, there remains a gap in comprehensively analyzing the factors contributing to the estimated dollar loss per fire incident in the Toronto area. To address this gap, we employ a data-driven approach, utilizing statistical modeling techniques to uncover patterns and relationships within the dataset. Our methodology involves exploring various factors such as property use, possible causes of the incident (e.g., mechanical/electrical failures), and the presence and functionality of fire alarms. Additionally, we incorporate the number of responding personnel from the Toronto Fire Services (TFS) as a variable to assess its impact on the estimated dollar loss.

The findings of this analysis hold significant implications for fire safety practices and policies in the Toronto area. By understanding the factors influencing the estimated dollar loss per fire incident, stakeholders can prioritize resources and implement targeted interventions to mitigate fire risks effectively. This paper is structured as follows: we first provide an overview of the dataset and variables considered, followed by the methodology employed for analysis. Subsequently, we present our findings, discussing the relationships uncovered and their implications for fire safety. Finally, we conclude by highlighting the importance of our findings and suggesting avenues for future research in this domain.

# Data {#sec-data}

All data collection and analysis was completed using statistical tools such as R [@citeR], and Rstudio IDE [@citeRStudio] to better streamline workflow. The analysis was done using the R program and the following supporting packages: `tidyverse` [@citeTidyverse], `dplyr` [@citeDplyr], `here` [@citeHere], `opendatatoronto` [@citeOpenDataToronto] and `ggplot2` [@citeGgplot2]. Further details regarding data cleaning and analysis will be discussed in the following subsections.

# Model {#sec-model}

Work in progress.

## Model set-up

Work in progress.

### Model justification

Work in progress.

# Results {#sec-results}

Work in progress.

# Discussion {#sec-discussion}

## First discussion point {#sec-first-point}

Work in progress.

## Second discussion point

Work in progress.

## Third discussion point

Work in progress.

## Weaknesses and next steps

Work in progress.

\newpage

\appendix

# Appendix {-}

Work in progress.

# Additional data details

Work in progress.

# Model details {#sec-model-details}

Work in progress.

## Posterior predictive check

Work in progress.

## Diagnostics

Work in progress.

\newpage

# References

### 2024-04-02 paper/paper.qmd
updated introduction section of paper

@@
 Developing a better understanding of the factors that influence the amount in damages per fire incident is crucial for implementing effective fire prevention measures to minimize the impact of such incidents.
-In this analysis, we delve into the extensive dataset provided by Toronto Open Data [@citeTorontoFireIncidents] which captures fire incidents within the Toronto area as defined by Ontario Fire Marshal (OFM) from 2011-2022.
+In this analysis, we delve into the extensive dataset provided by Toronto Open Data [@citeTorontoFireIncidents] which captures fire incidents within the Toronto area as defined by Ontario Fire Marshal (OFM) from 2011 to 2022.
 Through this dataset, we aim to predict the estimated dollar loss per fire incident across different property uses and identify the factors contributing to these losses.
 
-The motivation behind this study stems from the necessity to comprehend the underlying patterns and determinants of fire incident costs.
-By dissecting the data provided by Toronto Open Data, we seek to not only predict the estimated dollar loss but also to identify significant factors influencing these losses.
-Our analysis extends beyond mere prediction, delving into factors such as Toronto Fire Services' response time, area of origin, and the presence and functionality of fire alarms.
+The motivation behind this analysis is that fire safety matters and is important.
+Beyond the protection of assets, it's about keeping people safe.
+In order to ensure the safety of fellow Canadians, it is imperative to comprehend the underlying patterns and determinants of fire incident.
+By dissecting the data provided by Toronto Open Data, we seek to not only predict the estimated amount in damages but also to identify significant factors influencing these losses.
+Analyzing factors such as Toronto Fire Services' response time, area of origin, and the presence, functionality, and configuration of fire alarms.
 These additional variables are essential in understanding the dynamics of fire incidents and formulating informed strategies for fire prevention and mitigation.
@@
 To address this gap, we employ a data-driven approach, utilizing statistical modeling techniques to uncover patterns and relationships within the dataset.
-Our methodology involves exploring various factors such as property use, possible causes of the incident (e.g., mechanical/electrical failures), and the presence and functionality of fire alarms.
+Our methodology involves exploring various factors such as property use, possible causes of the incident (e.g., mechanical/electrical failures), and the presence of functioning fire alarms.
 Additionally, we incorporate the number of responding personnel from the Toronto Fire Services (TFS) as a variable to assess its impact on the estimated dollar loss.
@@
 The findings of this analysis hold significant implications for fire safety practices and policies in the Toronto area.
-By understanding the factors influencing the estimated dollar loss per fire incident, stakeholders can prioritize resources and implement targeted interventions to mitigate fire risks effectively.
-This paper is structured as follows: we first provide an overview of the dataset and variables considered, followed by the methodology employed for analysis.
-Subsequently, we present our findings, discussing the relationships uncovered and their implications for fire safety.
-Finally, we conclude by highlighting the importance of our findings and suggesting avenues for future research in this domain.
+By understanding the factors influencing the estimated dollar loss per fire incident, stakeholders can prioritize resources and allocate them accordingly to mitigate fire risks effectively.
+This paper is structured as follows: we first provide an overview of the dataset and variables of interest, followed by the methodology employed for analysis in @sec-data.
+Subsequently, we present our findings, discussing the relationships uncovered and their implications for fire safety in @sec-results.
+Finally, we conclude by highlighting the importance of our findings and suggesting avenues for future research in this domain concerning fire safety in @sec-discussion.
 

### 2024-04-16 other/datasheet/datasheet_fire_incidents.qmd
update datasheet

@@
 date-format: long
-abstract: "Work in progress."
 format: pdf
@@
 
-Extract of the questions from @gebru2021datasheets.
+Extract of the questions from Gebru et al. (2021).
 
@@
 1. *For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.*
-    - TBD
+    - The dataset provides more detailed information compared to the basic dataset that the Toronto Fire Service (TFS) responds to.
+The dataset was acquired through Toronto's Open Data which is an open source delivery tool that aims to connects people and data.
+These publicly available datasets are targeted towards various demographics such as app developers, students, or citizens.
 2. *Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?*
-    - TBD
+    - Toronto's Open Data, an open data initiative by the City of Toronto government.
 3. *Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.*
-    - TBD
+    - No funding was received.
 4. *Any other comments?*
-    - TBD
+    - No.
 
@@
 Please provide a description.*
-	- TBD
+	- Each row of the dataset is an individual instance of a fire incident that is recorded. Each column represents a specific point of information for each incident.
 2. *How many instances are there in total (of each type, if appropriate)?*
-	- TBD
+	- 29425
 3.
@@
 If it is not representative of the larger set, please describe why not (for example, to cover a more diverse range of instances, because instances were withheld or unavailable).*
-	- TBD
+	- All individuals included in the dataset are reported by the Ontario Fire Marshal (OFM) up to December 31, 2022.
+Incidents in the dataset that include incomplete data may be under investigation or is classified as a no loss outdoor fire.
+Some incidents have been excluded pursuant to exemptions under Section 8 of Municipal Freedom of Information and Protection of Privacy Act (MFIPPA).
+For privacy purposes, all personal information regarding the fire incidents have not been included in the dataset.
 4. *What data does each instance consist of? "Raw" data (for example, unprocessed text or images) or features? In either case, please provide a description.*
-	- TBD
+	- Each instance consists of information regarding the fire incident such as the time that the TFS received an emergency call to the time the respondents have arrived at the scene, estimated loss in damages, the classified source of the incident, etc.
+Various classifications for each incident was used throughout the dataset
 5. *Is there a label or target associated with each instance? If so, please provide a description.*
-	- TBD
+	- Yes, there is a unique id associated with each instance. This id is a unique number that represents the incident in the dataset.
 6.
@@
 This does not include intentionally removed information, but might include, for example, redacted text.*
-	- TBD
+	- Yes, there is missing information throughout the raw dataset.
+Information such as area of origin, building status, building impact, estimated dollars loss due to damages, and more.
+This information is likely unavailable due to ongoing investigations or information that was never recorded or received.
 7. *Are relationships between individual instances made explicit (for example, users' movie ratings, social network links)? If so, please describe how these relationships are made explicit.*
-	- TBD
+	- No, ther eare no relationships between instances.
 8. *Are there recommended data splits (for example, training, development/validation, testing)? If so, please provide a description of these splits, explaining the rationale behind them.*
-	- TBD
+	- No.
 9. *Are there any errors, sources of noise, or redundancies in the dataset? If so, please provide a description.*
-	- TBD
+	- No.
 10.
@@
 Please provide descriptions of all external resources and any restrictions associated with them, as well as links or other access points, as appropriate.*
-	- TBD
+	- Self-contained.
 11.
@@
 If so, please provide a description.*
-	- TBD
+	- No.
 12. *Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety? If so, please describe why.*
-	- TBD
+	- No.
 13.
@@
 If so, please describe how these subpopulations are identified and provide a description of their respective distributions within the dataset.*
-	- TBD
+	- No.
 14.
@@
 If so, please describe how.*
-	- TBD
+	- No.
 15.
@@
 If so, please provide a description.*
-	- TBD
+	- No.
 16. *Any other comments?*
-	- TBD
+	- No.
 
@@
 If so, please describe how.*
-	- TBD
+	- The data was gathered by the TFS throughout the occurrence of the incident, first beginning with a emergency phone call or report of a fire incident.
+Information such as estimated loss in damages is likely reported by the victims of the incident.
+A collaborative effort to collect information throughout the process.
 2.
@@
 How were these mechanisms or procedures validated?*
-	- TBD
+	- Telephone. Citizen reports and trained operators at the TFS.
 3. *If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?*
-	- TBD
+	- The dataset is not a sample from a larger set.
 4. *Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?*
-	- TBD
+	- Emergency call operators, firefighters, and citizens.
+Emergency dispatchers and firefighters classify as government employees and have their salaries publicly available by the Canadian government.
+Citizens that provide information are voluntary and do not receive monetary gain.
 5.
@@
 If not, please describe the timeframe in which the data associated with the instances was created.*
-	- TBD
+	- The dataset contains fire incidents from 2011 to 2022 which accurately represents the time frame of the data.
 6.
@@
 If so, please provide a description of these review processes, including the outcomes, as well as a link or other access point to any supporting documentation.*
-	- TBD
+	- No.
 7. *Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?*
-	- TBD
+	- The data was obtained from Toronto's Open Data which was published by Toronto Fire Services.
 8.
@@
 If so, please describe (or show with screenshots or other information) how notice was provided, and provide a link or other access point to, or otherwise reproduce, the exact language of the notification itself.*
-	- TBD
+	- Information is recorded by the government and associated government services.
+All 911 emergency calls are recorded for government records and all information protected through the MFIPPA (Municipal Freedom of Information and Protection of Privacy Act) will not be disclosed.
 9.
@@
 If so, please describe (or show with screenshots or other information) how consent was requested and provided, and provide a link or other access point to, or otherwise reproduce, the exact language to which the individuals consented.*
-	- TBD
+	- All individuals are subject to the MFIPPA and have the information concerning the incident recorded for government records.
 10.
@@
 If so, please provide a description, as well as a link or other access point to the mechanism (if appropriate).*
-	- TBD
+	- Information collected has revoked personal information for privacy purposes. Government records are subject to the MFIPPA.
 11.
@@
 If so, please provide a description of this analysis, including the outcomes, as well as a link or other access point to any supporting documentation.*
-	- TBD
+	- No.
 12. *Any other comments?*
-	- TBD
+	- No.
 
@@
 If not, you may skip the remaining questions in this section.*
-	- TBD
+	- No, no cleaning has been stated.
 2.
@@
 If so, please provide a link or other access point to the "raw" data.*
-	- TBD
+	- No.
 3. *Is the software that was used to preprocess/clean/label the data available? If so, please provide a link or other access point.*
-	- TBD
+	- None that is disclosed.
 4. *Any other comments?*
-	- TBD
+	- No.
 
@@
 1. *Has the dataset been used for any tasks already? If so, please provide a description.*
-	- TBD
+	- No.
 2. *Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.*
-	- TBD
+	- No.
 3. *What (other) tasks could the dataset be used for?*
-	- TBD
+	- General fire prevention measures.
 4.
@@
 Is there anything a dataset consumer could do to mitigate these risks or harms?*
-	- TBD
+	- No.
 5. *Are there tasks for which the dataset should not be used? If so, please provide a description.*
-	- TBD
+	- No.
 6. *Any other comments?*
-	- TBD
+	- No.
 
@@
 If so, please provide a description.*
-	- TBD
+	- No.
 2. *How will the dataset be distributed (for example, tarball on website, API, GitHub)? Does the dataset have a digital object identifier (DOI)?*
-	- TBD
+	- The dataset is publicly available at the Open Data portal. It can be downloaded directly, through API request, or the `opendatatoronto` R package.
 3. *When will the dataset be distributed?*
-	- TBD
+	- The dataset is available now.
 4.
@@
 If so, please describe this license and/ or ToU, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms or ToU, as well as any fees associated with these restrictions.*
-	- TBD
+	- The dataset is subject to Ontario's Open Government Licence outlined in the Open Data website.
 5.
@@
 If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.*
-	- TBD
+	- No.
 6.
@@
 If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any supporting documentation.*
-	- TBD
+	- No.
 7. *Any other comments?*
-	- TBD
+	- No.
 
@@
 1. *Who will be supporting/hosting/maintaining the dataset?*
-	- TBD
+	- The dataset is maintained by Toronto's Open Data and the TFS.
 2. *How can the owner/curator/manager of the dataset be contacted (for example, email address)?*
-	- TBD
+	- The publisher of the dataset can be contacted at the provided email address.
 3. *Is there an erratum? If so, please provide a link or other access point.*
-	- TBD
+	- No.
 4.
@@
 If so, please describe how often, by whom, and how updates will be communicated to dataset consumers (for example, mailing list, GitHub)?*
-	- TBD
+	- Yes, the dataset is stated to refresh annually. It appears as through the data is refreshed daily automatically.
 5.
@@
 If so, please describe these limits and explain how they will be enforced.*
-	- TBD
+	- No.
 6. *Will older versions of the dataset continue to be supported/hosted/maintained? If so, please describe how. If not, please describe how its obsolescence will be communicated to dataset consumers.*
-	- TBD
+	- No.
 7.
@@
 If so, please provide a description.*
-	- TBD
+	- No.
 8. *Any other comments?*
-	- TBD
+	- No.
 

### 2024-04-16 scripts/00-simulate_data.R
update data simulation script

@@ -10,6 +10,26 @@
 library(tidyverse)
 
 #### Simulate data ####
-# TODO set seed for simulation
+# set seed for simulation
+set.seed(905)
 
-# TODO create a simulation dataset with variables of interest following the proper distribution
+# create a simulation dataset with variables of interest following the proper distribution
+property_uses <- c("condo", "apartment", "single_detached", "warehouse") %>%
+  rep(each = 5, length.out = 20)
+
+estimated_dollar_loss <- rnorm(property_uses, 25000, 12500) %>%
+  round() %>%
+  pmax(0)
+
+estimated_loss_property_use <-
+  tibble(
+    property_use = property_uses,
+    estimated_dollar_loss = estimated_dollar_loss
+  )
+
+
+#### Test data ####
+## test data types##
+# methods used: https://stackoverflow.com/a/29591221
+all(sapply(estimated_loss_property_use$property_use, is.character))
+all(sapply(estimated_loss_property_use$estimated_dollar_loss, is.numeric))

### 2024-04-16 scripts/00-simulate_data.R
update simulation data

@@ -13,7 +13,7 @@ library(tidyverse)
 # set seed for simulation
 set.seed(905)
 
-# create a simulation dataset with variables of interest following the proper distribution
+# simulate estimated dollar loss and property use data
 property_uses <- c("condo", "apartment", "single_detached", "warehouse") %>%
   rep(each = 5, length.out = 20)
 
@@ -27,9 +27,26 @@ estimated_loss_property_use <-
     estimated_dollar_loss = estimated_dollar_loss
   )
 
+# simulate estimated dollar loss and area of origin data
+areas_of_origin <- c("kitchen", "trash", "office", "porch") %>%
+  rep(each = 5, length.out = 20)
+
+estimated_dollar_loss <- rnorm(areas_of_origin, 25000, 12500) %>%
+  round() %>%
+  pmax(0)
+
+estimated_loss_area_of_origin <-
+  tibble(
+    area_of_origin = areas_of_origin,
+    estimated_dollar_loss = estimated_dollar_loss
+  )
+
 
 #### Test data ####
 ## test data types##
 # methods used: https://stackoverflow.com/a/29591221
 all(sapply(estimated_loss_property_use$property_use, is.character))
 all(sapply(estimated_loss_property_use$estimated_dollar_loss, is.numeric))
+
+all(sapply(estimated_loss_area_of_origin$area_of_origin, is.character))
+all(sapply(estimated_loss_area_of_origin$estimated_dollar_loss, is.numeric))

### 2024-04-16 scripts/02-data_cleaning.R
update data cleaning script and data files

@@ -18,7 +18,7 @@ raw_fire_incidents_data <-
     file=here("data/raw_data/raw_fire_incidents.csv")
   )
 
-# clean the names and select column of interest
+# clean the names and select column of interest, 29425 entries
 cleaned_fire_incidents_data <-
   clean_names(raw_fire_incidents_data) |>
   select(
@@ -37,7 +37,21 @@ cleaned_fire_incidents_data <-
     tfs_arrival_time
   )
 
-# TODO clean selected variables of interest (omit NA values)
+# clean selected variables of interest (omit NA values), 15628 entries
+cleaned_fire_incidents_data <- cleaned_fire_incidents_data %>%
+  drop_na(area_of_origin,
+          estimated_dollar_loss,
+          latitude,
+          longitude,
+          method_of_fire_control,
+          number_of_responding_personnel,
+          possible_cause,
+          property_use,
+          smoke_alarm_at_fire_origin,
+          smoke_alarm_at_fire_origin_alarm_failure,
+          smoke_alarm_at_fire_origin_alarm_type,
+          tfs_alarm_time,
+          tfs_arrival_time)
 
 #### Save data ####
 # write cleaned data as csv

### 2024-04-16 scripts/03-test_data.R
update data cleaning script

@@ -9,8 +9,52 @@
 
 #### Workspace setup ####
 library(tidyverse)
+library(here)
+
+cleaned_fire_incidents_data <-
+  read_csv(
+    file=here("data/analysis_data/cleaned_fire_incidents.csv")
+  )
 
 #### Test data ####
-# TODO test data types for variables of interest
+## test data types##
+# methods used: https://stackoverflow.com/a/29591221
+all(sapply(cleaned_fire_incidents_data$area_of_origin, is.character))
+all(sapply(cleaned_fire_incidents_data$estimated_dollar_loss, is.numeric))
 
-# TODO test data values for variables of interest
+all(sapply(cleaned_fire_incidents_data$latitude, is.numeric))
+all(sapply(cleaned_fire_incidents_data$longitude, is.numeric))
+
+all(sapply(cleaned_fire_incidents_data$method_of_fire_control, is.character))
+all(sapply(cleaned_fire_incidents_data$number_of_responding_personnel, is.numeric))
+all(sapply(cleaned_fire_incidents_data$possible_cause, is.character))
+all(sapply(cleaned_fire_incidents_data$property_use, is.character))
+
+all(sapply(cleaned_fire_incidents_data$smoke_alarm_at_fire_origin, is.character))
+all(sapply(cleaned_fire_incidents_data$smoke_alarm_at_fire_origin_alarm_failure, is.character))
+all(sapply(cleaned_fire_incidents_data$smoke_alarm_at_fire_origin_alarm_type, is.character))
+
+all(sapply(cleaned_fire_incidents_data$tfs_alarm_time, is.double))
+all(sapply(cleaned_fire_incidents_data$tfs_arrival_time, is.double))
+
+## test data values ##
+all(cleaned_fire_incidents_data$area_of_origin != '')
+all(cleaned_fire_incidents_data$estimated_dollar_loss >= 0) # cannot have negative loss
+
+all(cleaned_fire_incidents_data$latitude >= -90 & cleaned_fire_incidents_data$latitude <= 90)
+all(cleaned_fire_incidents_data$longitude >= -180 & cleaned_fire_incidents_data$longitude <= 180)
+
+all(cleaned_fire_incidents_data$method_of_fire_control != '')
+all(cleaned_fire_incidents_data$number_of_responding_personnel >= 0)
+all(cleaned_fire_incidents_data$possible_cause != '')
[…]

### 2024-04-16 other/datasheet/datasheet_fire_incidents.qmd
update README and datasheet reference list

@@
 number-sections: true
+bibliography: references.bib
 ---
 
-Extract of the questions from Gebru et al. (2021).
+Extract of the questions from @gebru2021datasheets.
 

### 2024-04-16 scripts/02-data_cleaning.R
update data cleaning to create parquet file

@@ -10,6 +10,7 @@
 library(tidyverse)
 library(janitor)
 library(here)
+library(arrow)
 
 #### Clean data ####
 # load in raw data
@@ -58,4 +59,9 @@ cleaned_fire_incidents_data <- cleaned_fire_incidents_data %>%
 write_csv(
   cleaned_fire_incidents_data,
   "data/analysis_data/cleaned_fire_incidents.csv"
+)
+
+write_parquet(
+  cleaned_fire_incidents_data,
+  "data/analysis_data/cleaned_fire_incidents.parquet"
 )

### 2024-04-16 scripts/02-data_cleaning.R
update data cleaning and testing to parquet format

@@ -55,7 +55,7 @@ cleaned_fire_incidents_data <- cleaned_fire_incidents_data %>%
           tfs_arrival_time)
 
 #### Save data ####
-# write cleaned data as csv
+# write cleaned data as csv and parquet
 write_csv(
   cleaned_fire_incidents_data,
   "data/analysis_data/cleaned_fire_incidents.csv"

### 2024-04-16 scripts/03-test_data.R

@@ -10,10 +10,11 @@
 #### Workspace setup ####
 library(tidyverse)
 library(here)
+library(arrow)
 
 cleaned_fire_incidents_data <-
-  read_csv(
-    file=here("data/analysis_data/cleaned_fire_incidents.csv")
+  read_parquet(
+    file=here("data/analysis_data/cleaned_fire_incidents.parquet")
   )
 
 #### Test data ####

### 2024-04-17 paper/paper.qmd
update paper with data visualizations and data cleaning script

@@
 library(ggplot2)
+library(ggmap)
 library(here)
+library(arrow)
+
+analysis_data <- read_parquet(here("data/analysis_data/cleaned_fire_incidents.parquet"))
 ```
@@
 Developing a better understanding of the factors that influence the amount in damages per fire incident is crucial for implementing effective fire prevention measures to minimize the impact of such incidents.
-In this analysis, we delve into the extensive dataset provided by Toronto Open Data [@citeTorontoFireIncidents] which captures fire incidents within the Toronto area as defined by Ontario Fire Marshal (OFM) from 2011 to 2022.
-Through this dataset, we aim to predict the estimated dollar loss per fire incident across different property uses and identify the factors contributing to these losses.
+In this analysis, I delve into the extensive dataset provided by Toronto Open Data [@citeTorontoFireIncidents] which captures fire incidents within the Toronto area as defined by Ontario Fire Marshal (OFM) from 2011 to 2022.
+Through this dataset, I aim to predict the estimated dollar loss per fire incident across different property uses and identify the factors contributing to these losses.
 
@@
 In order to ensure the safety of fellow Canadians, it is imperative to comprehend the underlying patterns and determinants of fire incident.
-By dissecting the data provided by Toronto Open Data, we seek to not only predict the estimated amount in damages but also to identify significant factors influencing these losses.
+By dissecting the data provided by Toronto Open Data, I seek to not only predict the estimated amount in damages but also to identify significant factors influencing these losses.
 Analyzing factors such as Toronto Fire Services' response time, area of origin, and the presence, functionality, and configuration of fire alarms.
@@
 Despite the abundance of data available, there remains a gap in comprehensively analyzing the factors contributing to the estimated dollar loss per fire incident in the Toronto area.
-To address this gap, we employ a data-driven approach, utilizing statistical modeling techniques to uncover patterns and relationships within the dataset.
+To address this gap, I employ a data-driven approach, utilizing statistical modeling techniques to uncover patterns and relationships within the dataset.
 Our methodology involves exploring various factors such as property use, possible causes of the incident (e.g., mechanical/electrical failures), and the presence of functioning fire alarms.
-Additionally, we incorporate the number of responding personnel from the Toronto Fire Services (TFS) as a variable to assess its impact on the estimated dollar loss.
+Additionally, I incorporate the number of responding personnel from the Toronto Fire Services (TFS) as a variable to assess its impact on the estimated dollar loss.
 
@@
 By understanding the factors influencing the estimated dollar loss per fire incident, stakeholders can prioritize resources and allocate them accordingly to mitigate fire risks effectively.
-This paper is structured as follows: we first provide an overview of the dataset and variables of interest, followed by the methodology employed for analysis in @sec-data.
-Subsequently, we present our findings, discussing the relationships uncovered and their implications for fire safety in @sec-results.
-Finally, we conclude by highlighting the importance of our findings and suggesting avenues for future research in this domain concerning fire safety in @sec-discussion.
+This paper is structured as follows: I first provide an overview of the dataset and variables of interest, followed by the methodology employed for analysis in sec-data.
+Subsequently, I present our findings, discussing the relationships uncovered and their implications for fire safety in sec-results.
+Finally, I conclude by highlighting the importance of our findings and suggesting avenues for future research in this domain concerning fire safety in sec-discussion.
 
-# Data {#sec-data}
+# Data
 
 All data collection and analysis was completed using statistical tools such as R [@citeR], and Rstudio IDE [@citeRStudio] to better streamline workflow.
-The analysis was done using the R program and the following supporting packages: `tidyverse` [@citeTidyverse], `dplyr` [@citeDplyr], `here` [@citeHere], `opendatatoronto` [@citeOpenDataToronto] and `ggplot2` [@citeGgplot2].
+The analysis was done using the R program and the following supporting packages: `tidyverse` [@citeTidyverse], `dplyr` [@citeDplyr], `here` [@citeHere], `opendatatoronto` [@citeOpenDataToronto], `ggplot2` [@citeGgplot2], `ggmap` [@citeGgmap], `lubridate` [@citeLubridate], and `scales` [@citeScales].
 Further details regarding data cleaning and analysis will be discussed in the following subsections.
 
-# Model {#sec-model}
+# Model
 
@@
 
-# Results {#sec-results}
+# Results
 
-Work in progress.
+```{r graph1, fig.cap="Number of Fire Incidents in Toronto by Year (2011-2022)"}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-1
 
-# Discussion {#sec-discussion}
+# bar plot of the number of fire incidents per year in Toronto from 2011 to 2022
+fig_1_data <- analysis_data %>%
+  mutate(year= year(tfs_alarm_time))
 
-## First discussion point {#sec-first-point}
+fig_1_data <- fig_1_data %>%
+  group_by(year) %>%
+  summarize(incident_count = n())
+
+ggplot(fig_1_data, aes(x = factor(year), y = incident_count)) +
+  geom_bar(stat = "identity", fill = "navyblue") +
+  geom_text(aes(label = incident_count), vjust = -0.5, size = 3) +
+  labs(x = "Year", y = "Number of Incidents") +
+  theme_minimal()
+```
+
+@fig-1 testing testing hello.
+
+```{r graph2, fig.cap="Average Estimated Dollar Loss by Group Area of Origin"}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-2
+
+# bar plot of average dollar loss by area of origin
+fig_2_data <- analysis_data %>%
+  group_by(area_of_origin) %>%
+  summarize(avg_dollar_loss = round(mean(estimated_dollar_loss, na.rm = TRUE), 0))
+
+ggplot(fig_2_data, aes(x = area_of_origin, y = avg_dollar_loss)) +
+  geom_bar(stat = "identity", fill = "navyblue") +
+  geom_text(aes(label = scales::dollar(avg_dollar_loss)), vjust = -0.5, size = 3) +
+  labs(x = "Area of Origin", y = "Average Dollar Loss") +
+  scale_y_continuous(labels = scales::dollar) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+```
+
+@fig-2 testing testing hello
+
+```{r graph3, fig.cap="Fire Incidents Heatmap in Toronto"}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-3
+
+# heatmap of Toronto's fire incidents
+# https://docs.stadiamaps.com/tutorials/getting-started-in-r-with-ggmap/
+incident_location <- analysis_data %>%
+  select(
+    longitude,
+    latitude
+  )
+
+qmplot(longitude,
+       latitude,
+       data = incident_location,
+       maptype = "stamen_toner_lite",
+       color = I("red"), size = I(1),
+       alpha = I(0.05))
+
+```
+
+@fig-3 testing testing hello
+
+# Discussion
+
+## First discussion point
 
@@
 
-# Model details {#sec-model-details}
+# Model details
 

### 2024-04-17 scripts/02-data_cleaning.R

@@ -54,6 +54,31 @@ cleaned_fire_incidents_data <- cleaned_fire_incidents_data %>%
           tfs_alarm_time,
           tfs_arrival_time)
 
+# a function to group area_of_origin into broader categories
+group_areas <- function(area) {
+  if (grepl("Living|Sleeping|Dining|Cooking|Washroom|Sauna|Laundry|Closet", area)) {
+    return("Living and Common Areas")
+  } else if (grepl("Storage|Locker|Garage|Trash|Basement|Elevator|Chimney|Mechanical|Ducting|Utility|Awning|Porch|Crawl Space|Concealed|Attic|Open Area|Court|Parking|Fuel|Deck", area)) {
+    return("Storage and Structural")
+  } else if (grepl("Engine|Running Gear|Electrical|Fuel|Operator|Passenger|Trunk|Cargo|Vehicle", area)) {
+    return("Vehicle Components")
+  } else if (grepl("Manufacturing|Assembly|Laboratory|Operating Room|Performance|Backstage", area)) {
+    return("Functional and Performance Areas")
+  } else if (grepl("Office|Sales, Showroom|Records", area)) {
+    return("Office and Sales")
+  } else if (grepl("Entranceway|Hallway|Stairway|Court, Atrium|Other Means of Egress", area)) {
+    return("Entrance and Exit")
+  } else if (grepl("Exterior Wall|Roof|Concealed|Wall|Structural|Attic|Open Area|Court|Patio|Terrace", area)) {
+    return("Outside Areas")
+  } else if (grepl("Multiple|Residential/Business|Other - unclassified|Undetermined|Under Investigation", area)) {
+    return("Other/Undetermined")
+  } else {
+    return("Other")
+  }
+}
+
+cleaned_fire_incidents_data$area_of_origin <- sapply(cleaned_fire_incidents_data$area_of_origin, group_areas)
+
 #### Save data ####
 # write cleaned data as csv and parquet
 write_csv(

### 2024-04-18 paper/paper.qmd
update paper with completed data, results, and discussion sections

@@
 
+\newpage
 # Introduction
@@
 Fires pose a significant threat to both residential and commercial properties, often resulting in substantial financial losses and, more importantly, endangering lives.
-Developing a better understanding of the factors that influence the amount in damages per fire incident is crucial for implementing effective fire prevention measures to minimize the impact of such incidents.
-In this analysis, I delve into the extensive dataset provided by Toronto Open Data [@citeTorontoFireIncidents] which captures fire incidents within the Toronto area as defined by Ontario Fire Marshal (OFM) from 2011 to 2022.
-Through this dataset, I aim to predict the estimated dollar loss per fire incident across different property uses and identify the factors contributing to these losses.
+Developing a better understanding of the factors that influence the damage of fire incidents is crucial for implementing effective fire prevention measures to minimize the impact of such incidents.
+In this analysis, I explore the dataset provided by Toronto Open Data [@citeTorontoFireIncidents] which captures fire incidents from 2011 to 2022 within the Toronto area as defined by the Ontario Fire Marshal (OFM).
+Through this dataset, I aim to identify contributing factors to fire incidents and their resulting financial losses.
 
-The motivation behind this analysis is that fire safety matters and is important.
+The motivation behind this analysis stems from the importance of fire safety.
 Beyond the protection of assets, it's about keeping people safe.
-In order to ensure the safety of fellow Canadians, it is imperative to comprehend the underlying patterns and determinants of fire incident.
-By dissecting the data provided by Toronto Open Data, I seek to not only predict the estimated amount in damages but also to identify significant factors influencing these losses.
-Analyzing factors such as Toronto Fire Services' response time, area of origin, and the presence, functionality, and configuration of fire alarms.
+To ensure the safety of fellow Canadians, it is imperative to comprehend the underlying patterns and determinants of fire incident.
+Analyzing factors such as the fire's area of origin, and the presence, functionality, and configuration of fire alarms.
 These additional variables are essential in understanding the dynamics of fire incidents and formulating informed strategies for fire prevention and mitigation.
 
-Despite the abundance of data available, there remains a gap in comprehensively analyzing the factors contributing to the estimated dollar loss per fire incident in the Toronto area.
-To address this gap, I employ a data-driven approach, utilizing statistical modeling techniques to uncover patterns and relationships within the dataset.
-Our methodology involves exploring various factors such as property use, possible causes of the incident (e.g., mechanical/electrical failures), and the presence of functioning fire alarms.
+Despite the abundance of data available, there remains a gap in analyzing the factors that contributing to the estimated dollar loss per fire incident in the Toronto area such as sociological factors.
+To address this gap, I employ a data-driven approach, utilizing explorative data analysis techniques to reveal patterns and relationships within the dataset.
+My methodology involves exploring various factors such as property use, possible causes of the incident (e.g., mechanical/electrical failures), and the presence of functioning fire alarms.
 Additionally, I incorporate the number of responding personnel from the Toronto Fire Services (TFS) as a variable to assess its impact on the estimated dollar loss.
@@
 The findings of this analysis hold significant implications for fire safety practices and policies in the Toronto area.
-By understanding the factors influencing the estimated dollar loss per fire incident, stakeholders can prioritize resources and allocate them accordingly to mitigate fire risks effectively.
-This paper is structured as follows: I first provide an overview of the dataset and variables of interest, followed by the methodology employed for analysis in sec-data.
-Subsequently, I present our findings, discussing the relationships uncovered and their implications for fire safety in sec-results.
-Finally, I conclude by highlighting the importance of our findings and suggesting avenues for future research in this domain concerning fire safety in sec-discussion.
 
-# Data
+By understanding the factors influencing fire incidents and their associated financial losses, stakeholders can prioritize resources and allocate them accordingly to mitigate fire risks effectively.
+This paper is structured as follows: I first provide an overview of the dataset and variables of interest, followed by the methodology employed for analysis in @sec-data.
+Subsequently, I present my findings, discussing the relationships uncovered and their implications for fire safety in @sec-results.
+Finally, I conclude by highlighting the importance of the results by discussing the current laws and regulations in place and suggest avenues for future research in the domain of fire safety in @sec-discussion.
+
+# Data {#sec-data}
 
 All data collection and analysis was completed using statistical tools such as R [@citeR], and Rstudio IDE [@citeRStudio] to better streamline workflow.
-The analysis was done using the R program and the following supporting packages: `tidyverse` [@citeTidyverse], `dplyr` [@citeDplyr], `here` [@citeHere], `opendatatoronto` [@citeOpenDataToronto], `ggplot2` [@citeGgplot2], `ggmap` [@citeGgmap], `lubridate` [@citeLubridate], and `scales` [@citeScales].
-Further details regarding data cleaning and analysis will be discussed in the following subsections.
+The analysis was done using the R program and the following supporting packages: `tidyverse` [@citeTidyverse], `dplyr` [@citeDplyr], `here` [@citeHere], `opendatatoronto` [@citeOpenDataToronto], `ggplot2` [@citeGgplot2], `ggmap` [@citeGgmap], `arrow` [@citeArrow], `lubridate` [@citeLubridate], and `scales` [@citeScales].
+Further details regarding data cleaning, variables of interest, and measurement methodologies will be discussed in the following subsections.
 
-# Model
+## Data Source and Collection
 
-Work in progress.
+The data for this paper was acquired through the R package `opendatatoronto` [@citeOpenDataToronto], which provides publicly available datasets to be freely used, reused, and redistributed.
+Specifically, the "Fire Incidents" dataset is used as the basis of this analysis as it contains information regarding fire incidents in the area of Toronto provided by the Toronto Fire Services (TFS).
+The dataset captures incidents from the beginning of the year 2011 to the end of 2022.
+It is important to note that the dataset only includes incidents as defined by the Ontario Fire Marshal [@citeOFM].
 
-## Model set-up
+## Variables of Interest
 
-Work in progress.
+The dataset acquired contains various data points for analysis, but I only concern myself with a select few.
+In particular, the incident's area of origin, the estimated dollars loss in damages, the longitude and latitude of the incident, the number of responding personnel from the TFS, the possible causes of the incident, the determined property use, the presence, functionality and configuration of fire alarms, and finally the TFS notification and arrival time to the incidents.
 
-### Model justification
+## Measurement
 
-Work in progress.
+The methodologies behind the dataset's measurement is meticulously organized to account for scalability.
+Throughout the dataset, there are classifications methods used in scenarios where many instances require categorization.
+This is a decision that made with the concept of scalability in mind as there are far too many descriptors in the real world to account such that a singular description sufficiently and accurately describes the events of each incident.
+For example, regarding the area of origin, the raw data directly from Toronto's Open Data contains 74 unique classifications for an area of origin.
+The authors of the dataset hold the responsibility to properly assign the classification of each incident.
+This may result in inaccurate classifications without a clearly defined outline of what differentiates one from another.
 
-# Results
+One difficulty with measuring data concerning fire incidents is the nature of fire; fire is destructive and has the capacity to reduce tangible data to ash.
+The dataset has many instances of undetermined data which is likely related to nature of the incident.
+Taking a look at the types of smoke alarms at the locations of these incidents, nearly a fifth of all incidents recorded from 2011 to 2022 in Toronto are undetermined.
+While there may be numerous justifications for such values, it is difficult to analyze the data without disregarding a large portion of data, potentially leading to inconsistencies in the analysis and measurement errors.
 
-```{r graph1, fig.cap="Number of Fire Incidents in Toronto by Year (2011-2022)"}
+The dataset's concern for privacy is a bit alarming in regards to the provided latitude and longitude data points.
+The dataset contains information regarding the incident's nearest intersection which is generally comprised of two streets.
+While this alone does not breach the privacy of the victims of the fire incident, the latitude and longitude measurements bring you directly to the location of the incident.
+This may not be a concern for those two reside within high-rise apartments, but when considering detached homes with unique home addresses, this does raise some privacy concerns.
+
+# Results {#sec-results}
+
+In @fig-1, we can see the total number of fire incidents that occur each year from 2011 to 2022 in Toronto.
+While there is not much to decipher, we find that the number is relatively similar each year for the past decade.
+While some may argue that fire incidents have not reduced or increased substantially each year, another way of interpreting these numbers is that fire incidents have maintained a stable and infrequent occurrence in Toronto.
+
+```{r graph1, fig.cap="Number of fire incidents in Toronto by Year (2011-2022)"}
 #| echo: false
@@
 
-@fig-1 testing testing hello.
-
-```{r graph2, fig.cap="Average Estimated Dollar Loss by Group Area of Origin"}
+The dataset captures fire incidents that occur and are reported for the Toronto Fire Services and @fig-2 depicts where these incidents are located geographically.
+We can observe a greater number of incidents taking place in the heart of Toronto.
+This is likely due to the higher concentration of people with a population density of 3,088 persons per square kilometer which is the densest urban area or population centre in North America [@citeCensus2021].
+Looking further away from the city center, we find a higher concentration of incidents in the form of vertical lines that follow major roadways.
+Simply looking at the raw fire incident data, we find vehicles to be the third highest contributing factor to the source of ignition when considering electrical and mechanical incidents.
+```{r graph2, fig.cap="Heatmap of Toronto's fire incidents (2011-2022)"}
 #| echo: false
@@
 #| label: fig-2
-
-# bar plot of average dollar loss by area of origin
-fig_2_data <- analysis_data %>%
-  group_by(area_of_origin) %>%
-  summarize(avg_dollar_loss = round(mean(estimated_dollar_loss, na.rm = TRUE), 0))
-
-ggplot(fig_2_data, aes(x = area_of_origin, y = avg_dollar_loss)) +
-  geom_bar(stat = "identity", fill = "navyblue") +
-  geom_text(aes(label = scales::dollar(avg_dollar_loss)), vjust = -0.5, size = 3) +
-  labs(x = "Area of Origin", y = "Average Dollar Loss") +
-  scale_y_continuous(labels = scales::dollar) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
-```
-
-@fig-2 testing testing hello
-
-```{r graph3, fig.cap="Fire Incidents Heatmap in Toronto"}
-#| echo: false
-#| warning: false
-#| message: false
-#| label: fig-3
 
@@
        color = I("red"), size = I(1),
-       alpha = I(0.05))
-
+       alpha = I(0.05)
+       )
 ```
 
-@fig-3 testing testing hello
+Looking at @fig-3, the plot illustrates the average dollar loss in damages from fire incidents across various categorized areas of origin.
+Each bar representing a distinct area of origin, with the height indicating the average amount in damages.
+The categories with the highest average amount in damages all share the fact they are all highly populated areas or require the utmost fire safety precaution.
+While the leading area of origin is "other/undetermined", it is largely because of the dataset was unable to classify the incident to one particular area hence the high estimated loss due to damages.
+The largest estimated dollar loss recorded in the dataset which is estimated at $50 million occurred on May 7, 2019 at the York Memorial Collegiate Institute [@citeLargestCostFire].
+This is one example of an incident that is categorized under "other/undetermined" which resulted in a high dollar loss.
+As there are far too many classifications for large buildings that may serve many purposes, it is likely the reason such area of origin is leading.
 
-# Discussion
+```{r graph3, fig.cap="Average estimated dollar loss by grouped area of origin", fig.asp=1.0}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-3
+
+# bar plot of average dollar loss by grouped area of origin
+fig_2_data <- analysis_data %>%
+  group_by(area_of_origin) %>%
+  summarize(avg_dollar_loss = round(mean(estimated_dollar_loss, na.rm = TRUE), 0))
+
+ggplot(fig_2_data, aes(x = area_of_origin, y = avg_dollar_loss)) +
+  geom_bar(stat = "identity", fill = "navyblue") +
+  geom_text(aes(label = scales::dollar(avg_dollar_loss)), vjust = -0.5, size = 3) +
+  labs(x = "Area of Origin", y = "Average Dollar Loss") +
+  scale_y_continuous(labels = scales::dollar) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+```
+
+@fig-4 shows the number of each type of smoke alarms found present during fire incidents in Toronto from 2011 to 2022.
+Each bar representing a specific type or configuration of smoke alarm, while the height corresponds to the number of fire incidents associated with that particular fire alarm.
+We can see wireless smoke alarms to be the lowest number of smoke alarms found at the location of the incidents with the remaining smoke alarm types being within a reasonable range of each other.
+No alarm/undetermined is the most frequent which is likely related to the nature of the emergency as fire alarms may have been destroyed as a result.
+```{r graph4, fig.cap="Number of each type of smoke alarm present during fire incidents in Toronto", fig.asp=0.9}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-4
+
+# bar plot of the number of each type of smoke alarm present during fire incidents
+fig_4_data <- analysis_data %>%
+  filter(smoke_alarm_at_fire_origin_alarm_type != "9 - Type undetermined") %>% # ignore undetermined values
+  group_by(smoke_alarm_at_fire_origin_alarm_type) %>%
+  summarize(count = n())
+
+ggplot(fig_4_data, aes(x = smoke_alarm_at_fire_origin_alarm_type, y = count)) +
+  geom_bar(stat = "identity", fill = "navyblue") +
+  geom_text(aes(label = count), vjust = -0.5, size = 3) +
+  labs(x = "Smoke Alarm Type", y = "Count") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+```
+
+# Discussion {#sec-discussion}
 
@@
 
-Work in progress.
-
-## Second discussion point
-
-Work in progress.
-
-## Third discussion point
-
-Work in progress.
+This analysis shows that while there are many contributing factors and places that fire may occur causing potentially life-altering financial loss or loss of life, there are only a few safety measures to be made that can mitigate such events from happening.
+The Government of Ontario outlines home fire safety information including precautionary measures, laws and regulations regarding fire safety that is imperative for any homeowner, landlord, and tenant to ensure the safety of those around you [@citeFireSafety].
+While it is required for every home to have a smoke alarm on every story, the City of Toronto recommends replacing the batteries of your smoke alarms every two years [@citeFireSafetyToronto].
+As most fatal fires occur at night when people are asleep, after seeing the types of fire alarms present at fire incidents in @fig-4, it is advised that fire alarms be interconnected which would trigger all connected alarms if one where to ever activate.
+Making such change would be an incentive for public safety and also promote fire safety for all of those who reside within multiple unit dwellings especially in the busy city of Toronto.
+There are many systems that are already in place as preventative measures for the threat of fires almost everywhere today.
+From our schools to the underground subways, fire safety is present, but it is everyone's responsibility to be aware and understand the necessary procedures in the event of a fire.
+While Toronto is renowned for its diversity, one thing most would agree with is that fire has the capacity to bring happiness, despair, or destruction.
+It is important to acknowledge fire safety as it saves lives, protects our belongings, and ensures the safety of our communities from the devastating impacts of fires.
 
@@
 
-Work in progress.
+This analysis could be improved in regards to the consideration of sociological perspectives concerning peoples feelings towards fire safety.
+I believe discussing the awareness and sentiments towards fire safety and the steps they have taken to ensure the well-being for themselves and those around them.
+Including some survey or statistical evidence through a sociological lens would greatly benefit the overarching message of fire safety.
+Additionally, the analysis of the Toronto Fire Services response time to emergency calls could have been analyzed to determine whether or not it has changed in recent years as Toronto continues to increase in population.
 

### 2024-04-18 scripts/02-data_cleaning.R

@@ -77,7 +77,30 @@ group_areas <- function(area) {
   }
 }
 
-cleaned_fire_incidents_data$area_of_origin <- sapply(cleaned_fire_incidents_data$area_of_origin, group_areas)
+cleaned_fire_incidents_data$area_of_origin <-
+  sapply(cleaned_fire_incidents_data$area_of_origin, group_areas)
+
+# a function to clean and shorten the smoke alarm type values
+clean_smoke_alarm_names <- function(name) {
+  name_map <- c(
+    "8 - Not applicable - no smoke alarm or presence undetermined" = "No Alarm/Undetermined",
+    "2 - Hardwired (standalone)" = "Hardwired",
+    "1 - Battery operated" = "Battery Operated",
+    "4 - Interconnected" = "Interconnected",
+    "3 - Wireless" = "Wireless"
+  )
+  # Use the name map to clean the names
+  cleaned_name <- name_map[name]
+  # If the name is not found in the map, return the original name
+  if (is.na(cleaned_name)) {
+    return(name)
+  } else {
+    return(cleaned_name)
+  }
+}
+
+cleaned_fire_incidents_data$smoke_alarm_at_fire_origin_alarm_type <-
+  sapply(cleaned_fire_incidents_data$smoke_alarm_at_fire_origin_alarm_type, clean_smoke_alarm_names)
 
 #### Save data ####
 # write cleaned data as csv and parquet

### 2024-04-18 paper/paper.qmd
update README and title

@@
 ---
-title: "Evaluating Toronto's Fire Incidents to Extrapolate Future Occurence Data"
+title: "Toronto's Fire Incidents and Contributing Factors: An Explorative Analysis"
 author: 
@@
 date-format: long
-abstract: "This paper uses publicly available data from [Toronto Open Data](https://open.toronto.ca/) to predict the estimated dollar loss per fire incident in the Toronto area by property use.
+abstract: "This paper uses publicly available data from [Toronto Open Data](https://open.toronto.ca/) to identify contributing factors to fire incidents and their resulting financial losses.
 Further analysis of Toronto Fire Services' response time, area of origin, and various fire prevention measures as considering factors in the estimation.

### 2024-04-18 paper/paper.qmd
update appendix weaknesses and datasheet

@@
 
+A weakness of this analysis is the integrity of the dataset after cleaning and processing which is outlined in @sec-appendix-a.
 This analysis could be improved in regards to the consideration of sociological perspectives concerning peoples feelings towards fire safety.
@@
 Additionally, the analysis of the Toronto Fire Services response time to emergency calls could have been analyzed to determine whether or not it has changed in recent years as Toronto continues to increase in population.
+In @fig-2 we see that the most densely populated locations of Toronto also have the most concentrated number of fire incidents and incorporating some data model to predict the estimated dollar loss in populated areas would inform those within these areas of the dangers of fire.
 
@@
 
-# Appendix {-}
+# Appendix A {#sec-appendix-a}
 
-Work in progress.
+## Additional data details
 
-# Additional data details
+The dataset acquired through Toronto's Open Data [@citeOpenDataToronto] was processed using `R` [@citeR] which omitted all "NA" values as once one was present in any instance in the data, it was common to find many "NA" values in data points of the respective instance.
+This decision reduced the dataset from a total of 29,425 instances of data to 15,628 instances.
+Additionally, the data point `area_of_origin` was reorganized into broader categories for analysis as there were initially 74 unique areas of origin in the dataset.
+Furthermore, in @fig-4 the variable of interest is `smoke_alarm_at_fire_origin_alarm_type` which looked at the types of fire alarms without accounting for "undetermined" types as it provided little insight towards the goal of the paper.
 
-Work in progress.
+## Datasheet
+Extract of the questions from @gebru2021datasheets.
 
-# Model details
+**Motivation**
 
-Work in progress.
+1. *For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.*
+    - The dataset provides more detailed information compared to the basic dataset that the Toronto Fire Service (TFS) responds to.
+The dataset was acquired through Toronto's Open Data which is an open source delivery tool that aims to connects people and data.
+These publicly available datasets are targeted towards various demographics such as app developers, students, or citizens.
+2. *Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?*
+    - Toronto's Open Data, an open data initiative by the City of Toronto government.
+3. *Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.*
+    - No funding was received.
+4. *Any other comments?*
+    - No.
 
-## Posterior predictive check
+**Composition**
 
-Work in progress.
+1.
+*What do the instances that comprise the dataset represent (for example, documents, photos, people, countries)?
+Are there multiple types of instances (for example, movies, users, and ratings; people and interactions between them; nodes and edges)?
+Please provide a description.*
+	- Each row of the dataset is an individual instance of a fire incident that is recorded. Each column represents a specific point of information for each incident.
+2. *How many instances are there in total (of each type, if appropriate)?*
+	- 29425
+3.
+*Does the dataset contain all possible instances or is it a sample (not necessarily random) of instances from a larger set?
+If the dataset is a sample, then what is the larger set?
+Is the sample representative of the larger set (for example, geographic coverage)?
+If so, please describe how this representativeness was validated/verified.
+If it is not representative of the larger set, please describe why not (for example, to cover a more diverse range of instances, because instances were withheld or unavailable).*
+	- All individuals included in the dataset are reported by the Ontario Fire Marshal (OFM) up to December 31, 2022.
+Incidents in the dataset that include incomplete data may be under investigation or is classified as a no loss outdoor fire.
+Some incidents have been excluded pursuant to exemptions under Section 8 of Municipal Freedom of Information and Protection of Privacy Act (MFIPPA).
+For privacy purposes, all personal information regarding the fire incidents have not been included in the dataset.
+4. *What data does each instance consist of? "Raw" data (for example, unprocessed text or images) or features? In either case, please provide a description.*
+	- Each instance consists of information regarding the fire incident such as the time that the TFS received an emergency call to the time the respondents have arrived at the scene, estimated loss in damages, the classified source of the incident, etc.
+Various classifications for each incident was used throughout the dataset
+5. *Is there a label or target associated with each instance? If so, please provide a description.*
+	- Yes, there is a unique id associated with each instance. This id is a unique number that represents the incident in the dataset.
+6.
+*Is any information missing from individual instances?
+If so, please provide a description, explaining why this information is missing (for example, because it was unavailable).
+This does not include intentionally removed information, but might include, for example, redacted text.*
+	- Yes, there is missing information throughout the raw dataset.
+Information such as area of origin, building status, building impact, estimated dollars loss due to damages, and more.
+This information is likely unavailable due to ongoing investigations or information that was never recorded or received.
+7. *Are relationships between individual instances made explicit (for example, users' movie ratings, social network links)? If so, please describe how these relationships are made explicit.*
+	- No, ther eare no relationships between instances.
+8. *Are there recommended data splits (for example, training, development/validation, testing)? If so, please provide a description of these splits, explaining the rationale behind them.*
+	- No.
+9. *Are there any errors, sources of noise, or redundancies in the dataset? If so, please provide a description.*
+	- No.
+10.
+*Is the dataset self-contained, or does it link to or otherwise rely on external resources (for example, websites, tweets, other datasets)?
+If it links to or relies on external resources, a) are there guarantees that they will exist, and remain constant, over time; b) are there official archival versions of the complete dataset (that is, including the external resources as they existed at the time the dataset was created); c) are there any restrictions (for example, licenses, fees) associated with any of the external resources that might apply to a dataset consumer?
+Please provide descriptions of all external resources and any restrictions associated with them, as well as links or other access points, as appropriate.*
+	- Self-contained.
+11.
+*Does the dataset contain data that might be considered confidential (for example, data that is protected by legal privilege or by doctor-patient confidentiality, data that includes the content of individuals' non-public communications)?
+If so, please provide a description.*
+	- No.
+12. *Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety? If so, please describe why.*
+	- No.
+13.
+*Does the dataset identify any sub-populations (for example, by age, gender)?
+If so, please describe how these subpopulations are identified and provide a description of their respective distributions within the dataset.*
+	- No.
+14.
+*Is it possible to identify individuals (that is, one or more natural persons), either directly or indirectly (that is, in combination with other data) from the dataset?
+If so, please describe how.*
+	- No.
+15.
+*Does the dataset contain data that might be considered sensitive in any way (for example, data that reveals race or ethnic origins, sexual orientations, religious beliefs, political opinions or union memberships, or locations; financial or health data; biometric or genetic data; forms of government identification, such as social security numbers; criminal history)?
+If so, please provide a description.*
+	- No.
+16. *Any other comments?*
+	- No.
 
-## Diagnostics
+**Collection process**
 
-Work in progress.
+1.
+*How was the data associated with each instance acquired?
+Was the data directly observable (for example, raw text, movie ratings), reported by subjects (for example, survey responses), or indirectly inferred/derived from other data (for example, part-of-speech tags, model-based guesses for age or language)?
+If the data was reported by subjects or indirectly inferred/derived from other data, was the data validated/verified?
+If so, please describe how.*
+	- The data was gathered by the TFS throughout the occurrence of the incident, first beginning with a emergency phone call or report of a fire incident.
+Information such as estimated loss in damages is likely reported by the victims of the incident.
+A collaborative effort to collect information throughout the process.
+2.
+*What mechanisms or procedures were used to collect the data (for example, hardware apparatuses or sensors, manual human curation, software programs, software APIs)?
+How were these mechanisms or procedures validated?*
+	- Telephone. Citizen reports and trained operators at the TFS.
+3. *If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?*
+	- The dataset is not a sample from a larger set.
+4. *Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?*
+	- Emergency call operators, firefighters, and citizens.
+Emergency dispatchers and firefighters classify as government employees and have their salaries publicly available by the Canadian government.
+Citizens that provide information are voluntary and do not receive monetary gain.
+5.
+*Over what timeframe was the data collected?
+Does this timeframe match the creation timeframe of the data associated with the instances (for example, recent crawl of old news articles)?
+If not, please describe the timeframe in which the data associated with the instances was created.*
+	- The dataset contains fire incidents from 2011 to 2022 which accurately represents the time frame of the data.
+6.
+*Were any ethical review processes conducted (for example, by an institutional review board)?
+If so, please provide a description of these review processes, including the outcomes, as well as a link or other access point to any supporting documentation.*
+	- No.
+7. *Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?*
+	- The data was obtained from Toronto's Open Data which was published by Toronto Fire Services.
+8.
+*Were the individuals in question notified about the data collection?
+If so, please describe (or show with screenshots or other information) how notice was provided, and provide a link or other access point to, or otherwise reproduce, the exact language of the notification itself.*
+	- Information is recorded by the government and associated government services.
+All 911 emergency calls are recorded for government records and all information protected through the MFIPPA (Municipal Freedom of Information and Protection of Privacy Act) will not be disclosed.
+9.
+*Did the individuals in question consent to the collection and use of their data?
+If so, please describe (or show with screenshots or other information) how consent was requested and provided, and provide a link or other access point to, or otherwise reproduce, the exact language to which the individuals consented.*
+	- All individuals are subject to the MFIPPA and have the information concerning the incident recorded for government records.
+10.
+*If consent was obtained, were the consenting individuals provided with a mechanism to revoke their consent in the future or for certain uses?
+If so, please provide a description, as well as a link or other access point to the mechanism (if appropriate).*
+	- Information collected has revoked personal information for privacy purposes. Government records are subject to the MFIPPA.
+11.
+*Has an analysis of the potential impact of the dataset and its use on data subjects (for example, a data protection impact analysis) been conducted?
+If so, please provide a description of this analysis, including the outcomes, as well as a link or other access point to any supporting documentation.*
+	- No.
+12. *Any other comments?*
+	- No.
+
+**Preprocessing/cleaning/labeling**
+
+1.
+*Was any preprocessing/cleaning/labeling of the data done (for example, discretization or bucketing, tokenization, part-of-speech tagging, SIFT feature extraction, removal of instances, processing of missing values)?
+If so, please provide a description.
+If not, you may skip the remaining questions in this section.*
+	- No, no cleaning has been stated.
+2.
+*Was the "raw" data saved in addition to the preprocessed/cleaned/labeled data (for example, to support unanticipated future uses)?
+If so, please provide a link or other access point to the "raw" data.*
+	- No.
+3. *Is the software that was used to preprocess/clean/label the data available? If so, please provide a link or other access point.*
+	- None that is disclosed.
+4. *Any other comments?*
+	- No.
+
+**Uses**
+
+1. *Has the dataset been used for any tasks already? If so, please provide a description.*
+	- No.
+2. *Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.*
+	- No.
+3. *What (other) tasks could the dataset be used for?*
+	- General fire prevention measures.
+4.
+*Is there anything about the composition of the dataset or the way it was collected and preprocessed/cleaned/labeled that might impact future uses?
+For example, is there anything that a dataset consumer might need to know to avoid uses that could result in unfair treatment of individuals or groups (for example, stereotyping, quality of service issues) or other risks or harms (for example, legal risks, financial harms)?
+If so, please provide a description.
+Is there anything a dataset consumer could do to mitigate these risks or harms?*
+	- No.
+5. *Are there tasks for which the dataset should not be used? If so, please provide a description.*
+	- No.
+6. *Any other comments?*
+	- No.
+
+**Distribution**
+
+1.
+*Will the dataset be distributed to third parties outside of the entity (for example, company, institution, organization) on behalf of which the dataset was created?
+If so, please provide a description.*
+	- No.
+2. *How will the dataset be distributed (for example, tarball on website, API, GitHub)? Does the dataset have a digital object identifier (DOI)?*
+	- The dataset is publicly available at the Open Data portal. It can be downloaded directly, through API request, or the `opendatatoronto` R package.
+3. *When will the dataset be distributed?*
+	- The dataset is available now.
+4.
+*Will the dataset be distributed under a copyright or other intellectual property (IP) license, and/or under applicable terms of use (ToU)?
+If so, please describe this license and/ or ToU, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms or ToU, as well as any fees associated with these restrictions.*
+	- The dataset is subject to Ontario's Open Government Licence outlined in the Open Data website.
+5.
+*Have any third parties imposed IP-based or other restrictions on the data associated with the instances?
+If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.*
+	- No.
+6.
+*Do any export controls or other regulatory restrictions apply to the dataset or to individual instances?
+If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any supporting documentation.*
+	- No.
+7. *Any other comments?*
+	- No.
+
+**Maintenance**
+
+1. *Who will be supporting/hosting/maintaining the dataset?*
+	- The dataset is maintained by Toronto's Open Data and the TFS.
+2. *How can the owner/curator/manager of the dataset be contacted (for example, email address)?*
+	- The publisher of the dataset can be contacted at the provided email address.
+3. *Is there an erratum? If so, please provide a link or other access point.*
+	- No.
+4.
+*Will the dataset be updated (for example, to correct labeling errors, add new instances, delete instances)?
+If so, please describe how often, by whom, and how updates will be communicated to dataset consumers (for example, mailing list, GitHub)?*
+	- Yes, the dataset is stated to refresh annually. It appears as through the data is refreshed daily automatically.
+5.
+*If the dataset relates to people, are there applicable limits on the retention of the data associated with the instances (for example, were the individuals in question told that their data would be retained for a fixed period of time and then deleted)?
+If so, please describe these limits and explain how they will be enforced.*
+	- No.
+6. *Will older versions of the dataset continue to be supported/hosted/maintained? If so, please describe how. If not, please describe how its obsolescence will be communicated to dataset consumers.*
+	- No.
+7.
+*If others want to extend/augment/build on/contribute to the dataset, is there a mechanism for them to do so?
+If so, please provide a description.
+Will these contributions be validated/verified?
+If so, please describe how.
+If not, why not?
+Is there a process for communicating/distributing these contributions to dataset consumers?
+If so, please provide a description.*
+	- No.
+8. *Any other comments?*
+	- No.
 

### 2024-04-18 paper/paper.qmd
update data test script and fix grammar mistakes

@@
 ---
-title: "Toronto's Fire Incidents and Contributing Factors: An Explorative Analysis"
+title: "Toronto's Fire Incidents and Contributing Factors: An Exploratory Data Analysis"
 author: 
@@
 Beyond the protection of assets, it's about keeping people safe.
-To ensure the safety of fellow Canadians, it is imperative to comprehend the underlying patterns and determinants of fire incident.
+To ensure the safety of fellow Canadians, it is imperative to comprehend the underlying patterns and determinants of fire incidents.
 Analyzing factors such as the fire's area of origin, and the presence, functionality, and configuration of fire alarms.
@@
 
-Despite the abundance of data available, there remains a gap in analyzing the factors that contributing to the estimated dollar loss per fire incident in the Toronto area such as sociological factors.
-To address this gap, I employ a data-driven approach, utilizing explorative data analysis techniques to reveal patterns and relationships within the dataset.
+Despite the abundance of data available, there remains a gap in analyzing the factors that contribute to the estimated dollar loss per fire incident in the Toronto area such as sociological factors.
+To address this gap, I employ a data-driven approach, utilizing exploratory data analysis techniques to reveal patterns and relationships within the dataset.
 My methodology involves exploring various factors such as property use, possible causes of the incident (e.g., mechanical/electrical failures), and the presence of functioning fire alarms.
@@
 Throughout the dataset, there are classifications methods used in scenarios where many instances require categorization.
-This is a decision that made with the concept of scalability in mind as there are far too many descriptors in the real world to account such that a singular description sufficiently and accurately describes the events of each incident.
+This is a decision that was made with the concept of scalability in mind as there are far too many descriptors in the real world to account such that a singular description sufficiently and accurately describes the events of each incident.
 For example, regarding the area of origin, the raw data directly from Toronto's Open Data contains 74 unique classifications for an area of origin.
@@
 One difficulty with measuring data concerning fire incidents is the nature of fire; fire is destructive and has the capacity to reduce tangible data to ash.
-The dataset has many instances of undetermined data which is likely related to nature of the incident.
+The dataset has many instances of undetermined data which is likely related to the nature of the incident.
 Taking a look at the types of smoke alarms at the locations of these incidents, nearly a fifth of all incidents recorded from 2011 to 2022 in Toronto are undetermined.
@@
 The dataset's concern for privacy is a bit alarming in regards to the provided latitude and longitude data points.
-The dataset contains information regarding the incident's nearest intersection which is generally comprised of two streets.
+The dataset contains information regarding the incident's nearest intersection which is generally composed of two streets.
 While this alone does not breach the privacy of the victims of the fire incident, the latitude and longitude measurements bring you directly to the location of the incident.
-This may not be a concern for those two reside within high-rise apartments, but when considering detached homes with unique home addresses, this does raise some privacy concerns.
+This may not be a concern for those who reside within high-rise apartments, but when considering detached homes with unique home addresses, this does raise some privacy concerns.
 
@@
 Looking at @fig-3, the plot illustrates the average dollar loss in damages from fire incidents across various categorized areas of origin.
-Each bar representing a distinct area of origin, with the height indicating the average amount in damages.
+Each bar represents a distinct area of origin, with the height indicating the average amount in damages.
 The categories with the highest average amount in damages all share the fact they are all highly populated areas or require the utmost fire safety precaution.
-While the leading area of origin is "other/undetermined", it is largely because of the dataset was unable to classify the incident to one particular area hence the high estimated loss due to damages.
-The largest estimated dollar loss recorded in the dataset which is estimated at $50 million occurred on May 7, 2019 at the York Memorial Collegiate Institute [@citeLargestCostFire].
+While the leading area of origin is "other/undetermined", it is largely because the dataset was unable to classify the incident to one particular area hence the high estimated loss due to damages.
+The largest estimated dollar loss recorded in the dataset, which is estimated at $50 million occurred on May 7, 2019 at the York Memorial Collegiate Institute [@citeLargestCostFire].
 This is one example of an incident that is categorized under "other/undetermined" which resulted in a high dollar loss.
@@
 @fig-4 shows the number of each type of smoke alarms found present during fire incidents in Toronto from 2011 to 2022.
-Each bar representing a specific type or configuration of smoke alarm, while the height corresponds to the number of fire incidents associated with that particular fire alarm.
+Each bar represents a specific type or configuration of smoke alarm, while the height corresponds to the number of fire incidents associated with that particular fire alarm.
 We can see wireless smoke alarms to be the lowest number of smoke alarms found at the location of the incidents with the remaining smoke alarm types being within a reasonable range of each other.
@@
 A weakness of this analysis is the integrity of the dataset after cleaning and processing which is outlined in @sec-appendix-a.
-This analysis could be improved in regards to the consideration of sociological perspectives concerning peoples feelings towards fire safety.
+This analysis could be improved in regards to the consideration of sociological perspectives concerning people's feelings towards fire safety.
 I believe discussing the awareness and sentiments towards fire safety and the steps they have taken to ensure the well-being for themselves and those around them.

### 2024-04-18 scripts/03-test_data.R

@@ -54,8 +54,10 @@ all(cleaned_fire_incidents_data$smoke_alarm_at_fire_origin != '')
 all(cleaned_fire_incidents_data$smoke_alarm_at_fire_origin_alarm_failure != '')
 all(cleaned_fire_incidents_data$smoke_alarm_at_fire_origin_alarm_type != '')
 
-min(cleaned_fire_incidents_data$tfs_alarm_time) > as.Date("2010/12/31")
-max(cleaned_fire_incidents_data$tfs_alarm_time) < as.Date("2023/01/01")
+min(cleaned_fire_incidents_data$tfs_alarm_time) > as.POSIXct("2010/12/31")
+max(cleaned_fire_incidents_data$tfs_alarm_time) < as.POSIXct("2023/01/01")
 
-min(cleaned_fire_incidents_data$tfs_arrival_time) > as.Date("2010/12/31")
-max(cleaned_fire_incidents_data$tfs_arrival_time) < as.Date("2023/01/01")
+
+min(cleaned_fire_incidents_data$tfs_arrival_time) > as.POSIXct("2010/12/31")
+max(cleaned_fire_incidents_data$tfs_arrival_time) < as.POSIXct("2023/01/01")
+
